Predicting Conflict Escalation

### 2023-05-04 docs/hes_final/article.qmd
Extend conclusions with "so what?"

Further investigation showed that while overall numeric metrics look similar and there seems to be a slight performance advantage in using country data in addition to text features, the Transformer-based text-only model generalized better across the globe. Since our goal was to build a global model capable of predicting conflict escalation in any part of the world, the generalizability is a significant reason to prefer Transformer architecture and resampled dataset.

@@

We think that all three approaches produced promising results and the model choice depends on the user's goals, performance metrics and available resources.

@@

In this work we explored the possibility of predicting future geo-political conflict escalation at a country level using a global 
collection of news publications. We explored different ways of aggregating the data and applied both statistical machine learning and 
deep learning models. The results suggest that text embeddings provide a strong enough signal to make such predictions. Different data 
sampling and modeling approaches produce comparable results based on common metrics. Transformer-based model generalizes better across 
geographies while statistical model provides a size/cost of deployment advantage.

@@

One of the potential research extensions of this work is to continue experimenting with data sampling and aggregation approaches. We used either monthly averages to create $512\times1$ vectors or article level samples to create $512\times50$ matrices in our dataset preparation. While we received some promising results, it is possible that the first approach makes the data too coarse and the second too noisy. Another alternative could be generating daily averages and combining them into $512\times31$ (largest number of days in a month) matrices. This could potentially be more sensitive to the signal than the monthly averages and less noisy than article level samples.

@@

Building an application with a user interface could be a natural extension of this work. We envision a dashboard where a user can view a world map where countries are heatmap colored according to the estimated probability of escalating conflict. Countries with highest risk can be flagged or shown to a user in a separate view. In addition to the visual part of this application it is important to ensure that the model is deployed and can receive new data input and make inference.

>>>>

Further investigation showed that while overall numeric metrics look similar, the Transformer-based text-only model generalized better 
across the globe. Since our goal was to build a global model capable of predicting conflict escalation in any part of the world, the 
generalizability could be a reason to prefer Transformer architecture and resampled dataset.

@@

We think that all three approaches produced promising results and the model choice depends on the user's goals, selected performance 
metrics and available resources.

@@

In this work we explored the possibility of predicting future geo-political conflict escalation at a country level using a global 
collection of news publications. We compared the usage of text-only data and data enriched with country labels and metadata. We 
experimented with different ways of aggregating the data and applied both statistical machine learning and deep learning models. The 
results suggest that text embeddings provide a strong enough signal to make such predictions. Different data 
sampling and modeling approaches produced comparable results and each of the proposed models performed best at some of the common
numetic metrics. Transformer-based model generalized better across geographies while statistical model provided a size/cost of 
deployment advantage.

While this work did not produce a single best model for conflict escalation predictions, it demonstrated the possibility of such 
predictions from primarily text data. We provided a clear definition of a binary conflict escalation target and described several 
approaches for data aggregation and modeling. We hope this work will contribute to both UNICEF operations and political science 
research. The performance results can serve as a benchmark for future research projects in this 
field. We outline some ideas for future work in the next section.

@@

One of the potential research extensions of this work is to continue experimenting with data sampling and aggregation approaches. There 
are many alternatives to the approaches described above, for instance generating daily averages and combining them into $512\times31$ 
(largest number of days in a month) matrices. This could potentially be more sensitive to the signal than the monthly averages and less 
noisy than article level samples.

@@

Building an application with a user interface could be a natural extension of this work. We envision a dashboard where a user can view a 
world map where countries are heatmap colored according to the estimated probability of escalating conflict. Countries with highest risk 
can be flagged or shown to a user in a separate view. In addition to the visual part of this application, model deployment for easy
inference is an important project.
